Functional traits complicate the picture of temporal biodiversity change in bird and mammal communities

### 2024-04-01 paper/GEB Submission/manuscript_GEB.tex
Reformat discussion

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevalant than the other across many
taxa, biomes, and realms. At the study-level, the majority group of
studies showed no species or functional trends, but other types of
change were manifest for multiple studies, including studies
experiencing changes in functional redundancy, increased species and
functional richness, or loss of functional richness. With no one
prevailing scenario of change, it will be critical to link change
scenarios to ecological context.
\end{abstract}

@@

\textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
(13). While \textcolor{blue}{loss of functional diversity} is frequently
cited as one of the most pressing concerns of the anthropocene (18--20),
\textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
(21, 22). Forecasts of functional loss range from negligible (23) to
dire (24, 25). And while some observed trends show significant
functional loss (26) others document no loss even in some of the most
heavily impacted communities (27--29). On paleoecological time scales,
\textcolor{blue}{the functional space} shows mixed responses to
environmental change and extinction events (30, 31), with significant
impacts of species extinctions on functional diversity in some taxa and
not others (32). For some time periods, functional diversity appears to
be maintained for substantial portions of geological time (33).
\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds}
(34, 35) \textcolor{blue}{and ray-finned fishes, sharks, and rays} (36).

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and recent computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity change through time. We focus on mammal and bird species as
\textcolor{blue}{a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.}While
examining trends in plants, invertebrates, and other vertebrate species
is of equal interest, trait data for those taxa raise additional
challenges such as limited and biased species coverage (37), a lack of
accepted species-level means, and differences in the types of traits
collected. To ensure comparability across taxa in trait type and data
quality we therefore focus on mammals and birds.
\textcolor{blue}{We include body mass, dietary, foraging and other behavioral traits}
that were intentionally selected to be representative of a species'
Eltonian niche, thereby summarizing the functional role they play in the
community (38). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series
\textcolor{blue}{collected from the literature and ongoing monitoring efforts}.
\textcolor{blue}{Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations.}
Time series represent full assemblages rather than populations of single
species (31), and all time series include abundance of observed species.
Following best practices for the database (39), studies with multiple
sample locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 31 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations.
\textcolor{blue}{The resulting cell size for our data was approximately 95 }km\textsuperscript{2}.
All samples from a study within a single cell were considered to be a
single time series, and species abundances were combined for all
samples.

We used trait data from the Elton Trait Database, which consists of
species-level means for traits that represent species' multifaceted role
in the community (38). Traits include: body mass, diet, nocturnality,
forest foraging strata, pelagic use. With the exception of body mass,
traits were broken down into percentage or binary use for each level of
the trait type (Table \ref{tab:traitTable}).

@@

\caption{\label{tab:traitTable}Description of the traits included in the analysis broken down by categories at data type.}
\centering
\begin{tabular}[t]{l|l|l}
\hline
Trait & Category & Data Type\\
\hline
 & Invertebrate & \\
\cline{2-2}
 & Mammals and Birds & \\
\cline{2-2}
 & Reptiles & \\
\cline{2-2}
 & Fish & \\
\cline{2-2}
 & Unknown Vertebrates & \\
\cline{2-2}
 & Scavenging & \\
\cline{2-2}
 & Fruit & \\
\cline{2-2}
 & Nectar & \\
\cline{2-2}
 & Seeds & \\
\cline{2-2}
\multirow{-10}{*}{\raggedright\arraybackslash Diet} & Other Plant & \multirow{-10}{*}{\raggedright\arraybackslash percentage consumed}\\
\cline{1-3}
 & Below water surface & \\
\cline{2-2}
 & water surface & \\
\cline{2-2}
 & ground & \\
\cline{2-2}
 & understory & \\
\cline{2-2}
 & > 2m, below canopy & \\
\cline{2-2}
 & canopy & \\
\cline{2-2}
\multirow{-7}{*}{\raggedright\arraybackslash Foraging Strata} & aerial & \multirow{-7}{*}{\raggedright\arraybackslash percentage of use}\\
\cline{1-3}
 & yes & \\
\cline{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Pelagic Specialist} & no & \\
\cline{1-2}
 & yes & \\
\cline{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Nocturnal} & no & \multirow{-4}{*}{\raggedright\arraybackslash binary}\\
\cline{1-3}
Body Mass & - & continous, in grams\\
\hline
\end{tabular}
\end{table}

In order to ensure taxonomic consistency across datasets, BioTIME
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (retrieved
09-15-2020 from the on-line database,
\url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
\texttt{taxadb} R package (40, 41). If more than one species in the
assemblage data resolved to the same identifier, observations were
considered the same species. For trait data, traits for all species of
the same identifier were averaged. Only studies for which at least 75\%
of species had trait data were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 5 species observed were also excluded. Sensitivity analyses
were conducted for the trait coverage threshold and the duration of
included time series.

@@

Functional diversity metrics were calculated using the \emph{dbFD}
function from the \emph{FD} R package (42). Here we report functional
richness (\emph{FRic}), functional evenness (\emph{FEve}), and
functional divergence (\emph{FDiv}) which together describe three
complementary characteristics of the functional space (43, 44).
\emph{FRic} assesses the volume of the trait space occupied by species
in the community, with higher values indicating communities with species
of more extreme trait values. \emph{FEve} describes how species are
distributed across the trait space and how abundance is distributed
across species. Higher values of \emph{FEve} indicate more even spacing
of species in the trait space and individuals across species.
\emph{FDiv} measures the degree to which species and their abundances
maximize differences in the functional space. Higher values of
\emph{FDiv} therefore correspond to communities where many highly
abundant species are on the edges of the trait space. We also calculated
the community-weighted mean (\emph{CWM}) of included traits to examine
shifts in the distribution of each trait.
\textcolor{blue}{Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (45, 46).
\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
the maximum number of traits that fulfills the criteria \(s >= 2^t\),
where \(s\) is the number of species and \emph{t} is the number of
traits. This restriction allows for enough axes to capture the trait
space while maintaining computational feasibility (47). Metrics
incorporated weighting based on species abundance.

@@

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
from null estimates (48). Null model corrections allow us to assess the
degree to which the observed functional diversity metric deviates from
the value expected by chance in a randomly assembled community. Null
estimates were calculated for each rarefied sample by randomly sampling
species from the species pool for each year and randomly assigning
observed abundances to species.
\textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
This process was repeated 500 times to get an estimate and standard
deviation of the null expectation for the metric for each rarefaction
sample for that time series. We used these values to calculate SES using
the following formula:
\(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We then
calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.
\textcolor{blue}{This approach will be less accurate for shorter time series, as we likley will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.}

@@

To test for trends within and between different levels of taxa, biome,
realm, and protection status we fit separate models with each of those
factors added as a predictor interacting with time to the original model
structure. We estimated within-level slopes and calculated between-level
contrasts using the \emph{emmeans} package (v1.8.2, 49).
\textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
We assessed the impact of time series duration and start year on
study-level trends using linear models with duration and start year as
predictors. All models in our analysis were fit using the \emph{lme4}
(v1.1-30) package in R (v4.2.3) and p-values were calculated by
Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(50--52).

@@

\hypertarget{evidence-of-consensus}{%
\subsubsection{Evidence of consensus}\label{evidence-of-consensus}}

\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities. We did not detect an overall trend in any functional diversity metric, corrected or uncorrected. As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time}
(31)\textcolor{blue}{, indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends. We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial}
(39).
\textcolor{blue}{However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.}

\textcolor{blue}{We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities. This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe}
(53--55).
\textcolor{blue}{However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories}
(53).
\textcolor{blue}{In general, functional capacity of mammals shows greater declines that bird communities.}

\textcolor{blue}{We also found evidence of increasing evenness for Marine studies. The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability }
(56),
\textcolor{blue}{as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change }(44,
57).
\textcolor{blue}{Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities }(58).
\textcolor{blue}{The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work }(59).

\textcolor{blue}{Compositional shifts shown by significant trends in CWM's were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass. Increases in fish consumption for bird and temperate studies my be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing. However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch. Decreases in nectar consumption for tropical studies is likely due declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data. Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers. Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.}

\hypertarget{study-heterogeneity}{%
\subsubsection{Study heterogeneity}\label{study-heterogeneity}}

@@

\textcolor{blue}{The lack of trend could be the result of multiple possible scenarios. First, these may be communities resisting perturbations or simply not experiencing significant perturbations. Given the studies in this group come from all possible taxa, realms, biomes, and protection statuses, evidence points to communities resisting perturbation, offering some hope that there are areas of the globe where communities are fairing well for now. Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window. This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago }(60).
\textcolor{blue}{Third, these communities may be experiencing directional shifts undetectable by available data. For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in the functional space as a whole and impact maintenance of ecological processes.}

\textcolor{blue}{Three of the groups fit under the broad umbrella of changes in redundancy. By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella. For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced }(61).
\textcolor{blue}{These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future }(62).
\textcolor{blue}{Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone}
(63).

\textcolor{blue}{The next group of studies (3 studies) exhibit an increase in species richness and functional richness with significant turnover. While these communities are losing some species that are functionally redundant or functional analogs of species additions, they are gaining even more species that expand the functional space. These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function. Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden). Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline}
(35)
\textcolor{blue}{, and loss of common, functionally general species even as rare species are increasing in North America and Europe }
(64--66).

\textcolor{blue}{The final group (3 studies) exhibits decreasing functional evenness. These communities are of particular concern not just because of the role evenness plays in maintaining functionality, but because of increasing evidence evenness is more sensitive to environmental change than species richness}
(56).
\textcolor{blue}{Formal tests of evenness changes as an early warning signal of more catastrophic shifts in functional diversity will be critical next step for understanding the full implication of evenness loss for ecosystem processes.}

>>>>

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevalant than the other across many
taxa, biomes, and realms. At the study-level, the majority group of
studies showed no species or functional trends, but other types of
change were found for multiple studies, including studies experiencing
changes in functional redundancy, increased species and functional
richness, or loss of functional richness. With no one prevailing
scenario of change, it will be critical to link change scenarios to
ecological context.
\end{abstract}

@@

\textcolor{blue}{The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow taxonomic trends}
(13). While \textcolor{blue}{loss of functional diversity} is frequently
cited as one of the most pressing concerns of the anthropocene (18--20),
\textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
(21, 22). Forecasts of functional loss range from negligible (23) to
dire (24, 25). And while some observed trends show significant
functional loss (26) others document no loss even in some of the most
heavily impacted communities (27--29). On paleoecological time scales,
\textcolor{blue}{the functional space} shows mixed responses to
environmental change and extinction events (30, 31), with significant
impacts of species extinctions on functional diversity in some taxa and
not others (32). For some time periods, functional diversity appears to
be maintained for substantial portions of geological time (33).
\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds}
(34, 35) \textcolor{blue}{and ray-finned fishes, sharks, and rays} (36).

We have stronger expectations for changes in the prevalence of some
individual traits. For example, animal body size is expected to decrease
as a result of climate change, a phenomena that has been documented in
multiple taxa empirically and experimentally (37--41). There is some
evidence that this holds true for birds (42), but the picture is likely
more complicated for mammals, where urbanization may actually lead to
larger body sizes due to novel food sources (43), even as megafaunal
loss leads to on average smaller community body size. For dietary
traits, recent work documenting insect declines (citation) points to
potentially significant negative impacts on insectivorous animals
(44--46). Predicted extinctions based on species-level vulnerability
point to further dietary shifts, favoring increases in invertivorous
species (47). Some systems also show significant shifts in the
prevalence of different kinds of foragers in birds, for example loss of
arboreal foragers in agricultural systems (48), and loss of neotropical
understory foragers even in protect areas (49).

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and recent computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity and composition change through time. We focus on mammal and
bird species as
\textcolor{blue}{a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.}While
examining trends in plants, invertebrates, and other vertebrate species
is of equal interest, trait data for those taxa raise additional
challenges such as limited and biased species coverage (50), a lack of
accepted species-level means, and differences in the types of traits
collected. To ensure comparability across taxa in trait type and data
quality we therefore focus on mammals and birds.
\textcolor{blue}{We include body mass, dietary, foraging and other behavioral traits}
that were intentionally selected to be representative of a species'
Eltonian niche, thereby summarizing the functional role they play in the
community (51). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. We present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}
We further evaluate evidence for key predictions of individual trait
prevalence from the literature, including changes in mean body size,
dietary traits, and foraging strata.`

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series
\textcolor{blue}{collected from the literature and ongoing monitoring efforts}.
\textcolor{blue}{Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations.}
Time series represent full assemblages rather than populations of single
species (31), and all time series include abundance of observed species.
Following best practices for the database (52), studies with multiple
sample locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 31 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations.
\textcolor{blue}{The resulting cell size for our data was approximately 95 }km\textsuperscript{2}.
All samples from a study within a single cell were considered to be a
single time series, and species abundances were combined for all
samples.

We used trait data from the Elton Trait Database, which consists of
species-level means for traits that represent species' multifaceted role
in the community (51). Traits include: body mass, diet, nocturnality,
forest foraging strata, pelagic use. With the exception of body mass,
traits were broken down into percentage or binary use for each level of
the trait type (Table \ref{tab:traitTable}).

@@

\caption{\label{tab:traitTable}Description of the traits included in the analysis broken down by categories at data type.}
\centering
\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{3cm}|>{\raggedright\arraybackslash}p{3cm}}
\hline
Trait & Category & Taxa & Data Type\\
\hline
 & Invertebrate &  & \\
\cline{2-2}
 & Mammals and Birds &  & \\
\cline{2-2}
 & Reptiles &  & \\
\cline{2-2}
 & Fish &  & \\
\cline{2-2}
 & Unknown Vertebrates &  & \\
\cline{2-2}
 & Scavenging &  & \\
\cline{2-2}
 & Fruit &  & \\
\cline{2-2}
 & Nectar &  & \\
\cline{2-2}
 & Seeds &  & \\
\cline{2-2}
\multirow{-10}{*}{\raggedright\arraybackslash Diet} & Other Plant & \multirow{-10}{*}{\raggedright\arraybackslash Bird and Mammal} & \multirow{-10}{*}{\raggedright\arraybackslash percentage consumed}\\
\cline{1-4}
 & Below water surface &  & \\
\cline{2-2}
 & water surface &  & \\
\cline{2-2}
 & ground &  & \\
\cline{2-2}
 & understory &  & \\
\cline{2-2}
 & > 2m, below canopy &  & \\
\cline{2-2}
 & canopy &  & \\
\cline{2-2}
\multirow{-7}{*}{\raggedright\arraybackslash Foraging Strata} & aerial &  & \multirow{-7}{*}{\raggedright\arraybackslash percentage of use}\\
\cline{1-2}
\cline{4-4}
 & yes &  & \\
\cline{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Pelagic Specialist} & no & \multirow{-9}{*}{\raggedright\arraybackslash Bird} & \\
\cline{1-3}
 & yes &  & \\
\cline{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Nocturnal} & no & \multirow{-2}{*}{\raggedright\arraybackslash Bird and Mammal} & \\
\cline{1-3}
 & yes &  & \\
\cline{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Crepuscular} & no &  & \\
\cline{1-2}
 & yes &  & \\
\cline{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Diurnal} & no & \multirow{-4}{*}{\raggedright\arraybackslash Mammal} & \multirow{-8}{*}{\raggedright\arraybackslash binary}\\
\cline{1-4}
Body Mass & - & Bird and Mammal & continous, in grams\\
\hline
\end{tabular}
\end{table}

In order to ensure taxonomic consistency across datasets, BioTIME
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (retrieved
09-15-2020 from the on-line database,
\url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
\texttt{taxadb} R package (53, 54). If more than one species in the
assemblage data resolved to the same identifier, observations were
considered the same species. For trait data, traits for all species of
the same identifier were averaged. Only studies for which at least 75\%
of species had trait data were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 5 species observed were also excluded. Sensitivity analyses
were conducted for the trait coverage threshold and the duration of
included time series.

@@

Functional diversity metrics were calculated using the \emph{dbFD}
function from the \emph{FD} R package (55). Here we report functional
richness (\emph{FRic}), functional evenness (\emph{FEve}), and
functional divergence (\emph{FDiv}) which together describe three
complementary characteristics of the functional space (56, 57).
\emph{FRic} assesses the volume of the trait space occupied by species
in the community, with higher values indicating communities with species
of more extreme trait values. \emph{FEve} describes how species are
distributed across the trait space and how abundance is distributed
across species. Higher values of \emph{FEve} indicate more even spacing
of species in the trait space and individuals across species.
\emph{FDiv} measures the degree to which species and their abundances
maximize differences in the functional space. Higher values of
\emph{FDiv} therefore correspond to communities where many highly
abundant species are on the edges of the trait space.

We also calculated the community-weighted mean (\emph{CWM}) of all
continuous traits to examine \textcolor{blue}{turnover} in the
distribution of each trait.
\textcolor{blue}{Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained. CWM's are therefore a way to assess whether or not that turnover is occurring and what the nature of the shift may be. Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (58, 59).
\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
the maximum number of traits that fulfills the criteria \(s >= 2^t\),
where \(s\) is the number of species and \emph{t} is the number of
traits. This restriction allows for enough axes to capture the trait
space while maintaining computational feasibility (60). Metrics
incorporated weighting based on species abundance.

@@

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
from null estimates (61). Null model corrections allow us to assess the
degree to which the observed functional diversity metric deviates from
the value expected by chance in a randomly assembled community. Null
estimates were calculated for each rarefied sample by randomly sampling
species from the species pool for each year and randomly assigning
observed abundances to species.
\textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
This process was repeated 500 times to get an estimate and standard
deviation of the null expectation for the metric for each rarefaction
sample for that time series. We used these values to calculate SES using
the following formula:
\(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We then
calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.
\textcolor{blue}{This approach will be less accurate for shorter time series, as we likley will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.}

@@

To test for trends within and between different levels of taxa, biome,
realm, and protection status we fit separate models with each of those
factors added as a predictor interacting with time to the original model
structure. We estimated within-level slopes and calculated between-level
contrasts using the \emph{emmeans} package (v1.8.2, 62).
\textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
We assessed the impact of time series duration and start year on
study-level trends using linear models with duration and start year as
predictors. All models in our analysis were fit using the \emph{lme4}
(v1.1-30) package in R (v4.2.3) and p-values were calculated by
Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(63--65).

@@

\hypertarget{evidence-of-consensus}{%
\subsection{Evidence of consensus}\label{evidence-of-consensus}}

\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities. We did not detect an overall trend in any functional diversity metric, corrected or uncorrected. As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time}
(31)\textcolor{blue}{, indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends. We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial}
(52).
\textcolor{blue}{However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.}

\textcolor{blue}{We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities. This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe}
(66--68).
\textcolor{blue}{However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories}
(66).
\textcolor{blue}{In general, functional capacity of mammals showed greater declines that bird communities.}

\hypertarget{areas-of-incongruence}{%
\subsection{Areas of incongruence}\label{areas-of-incongruence}}

\textcolor{blue}{The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or simply unexpected. For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation. The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas. Rather than the result of protection itself, this is likely due to the fact that the majority of studies were from terrestrial, temperate studies where turnover is known to be lower }(52).
\textcolor{blue}{While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts. Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.}

\textcolor{blue}{We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems. The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability }
(69),
\textcolor{blue}{as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change }(57,
70).
\textcolor{blue}{Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities }(71).
\textcolor{blue}{An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world. Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work }(72).

Our results were also inconsistent with all predictions for changes in
trait prevalence, including foraging strata, body size, and dietary
traits.
\textcolor{blue}{We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species. These shifts may therefore be the result of specific contexts, and not generalizable to bird communities across the globe. We found no evidence of general trends in the *CWM* for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data. Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing}
(73).
\textcolor{blue}{Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened}
(74).

Rather than declines in insectivory and increases in invertivory
predicted by previous work, we instead found changes in the degree of
consumption of fish, nectar, and vertebrates.
\textcolor{blue}{Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing. However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, which makes these results suprising. Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data. Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers. Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.}

\hypertarget{study-heterogeneity}{%
\subsection{Study heterogeneity}\label{study-heterogeneity}}

@@

\textcolor{blue}{The lack of trend could be the result of multiple possible scenarios. First, these may be communities resisting perturbations or simply not experiencing significant perturbations. Given the studies in this group come from all possible taxa, realms, biomes, and protection statuses, evidence points to communities resisting perturbation, offering some hope that there are areas of the globe where communities are fairing well for now. Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window. This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago }(75).
\textcolor{blue}{Third, these communities may be experiencing directional shifts undetectable by available data. For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in the functional space as a whole and impact maintenance of ecological processes.}

\textcolor{blue}{Three of the groups fit under the broad umbrella of changes in redundancy. By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella. For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced }(76).
\textcolor{blue}{These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future }(77).
\textcolor{blue}{Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone}
(78).

\textcolor{blue}{The next group of studies (3 studies) exhibit an increase in species richness and functional richness with significant turnover. While these communities are losing some species that are functionally redundant or functional analogs of species additions, they are gaining even more species that expand the functional space. These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function. Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden). Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline}
(35)
\textcolor{blue}{, and loss of common, functionally general species even as rare species are increasing in North America and Europe }
(79--81).

\textcolor{blue}{The final group (3 studies) exhibits decreasing functional evenness. These communities are of particular concern not just because of the role evenness plays in maintaining functionality, but because of increasing evidence evenness is more sensitive to environmental change than species richness}
(69).
\textcolor{blue}{Formal tests of evenness changes as an early warning signal of more catastrophic shifts in functional diversity will be critical next step for understanding the full implication of evenness loss for ecosystem processes.}
